GEE models

### 2017-07-20 doc/manuscript.Rmd
updated the results

For this study, there were 
`r female$npct` females and `r european$npct` had European-ancestry (see 
`r cite_t("basic")`). The mean follow-up time was `r calc_followup_time()`
years, where `r calc_pct_full_visits()`% of participants attended all three
visits.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes.
Insulin sensitivity and beta-cell function measures showed a significant median
decline from baseline to 6-years between
`r chg_outcomes$chg` (p`r chg_outcomes$p`, based on a time-adjusted GEE model)
in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and pre-diabetes (IFG and IGT),
respectively, over the 6-years. Incident diabetes cases were excluded from further GEE
analyses.

`r cite_f('nefa')` shows the compositional distribution of NEFA in the study
participants. Four individual NEFA made up the vast majority 
(`r sum(nefa$pct)`%) of the total NEFA fraction. The largest contributors
were `r nefa$c[1]`, `r nefa$c[2]`, `r nefa$c[3]`, and 
`r nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefa')`. 
There were no differences in individual NEFA for those who did compared to those
who did not take medications to manage blood lipids (data not shown).
All individual NEFA as well as the total fraction had
correlations that ranged from weak to null (r<0.3) with participant
characteristics (see `r cite_sf('corr_conc')` and `r cite_sf("corr_pct")`).

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
While full model adjustment (`r cite_f('gee_adj')`) attenuated most of these
associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
continued to show negative associations with IGI/IR and ISSI-2. Sensitivity
analyses revealed that adjustment for WC was responsible for attenuating the
unadjusted results. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD increase
in any of these three NEFA variables, there was an average predicted
`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at each
clinic visit. However, in contrast to these results using NEFA modeled as
concentrations, none of the NEFA variables modeled as mol% were associated with
the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
ISSI-2 (data not shown), though this may be due to the fact that only a subset
of participants at the 6 year visit had TAG measured. Sensitivity analyses to
examine the association with
the sum of fatty acid classes (saturated vs unsaturated) revealed similar
findings as seen with the adjusted GEE model for total NEFA (data not shown).
There were no significant interaction effects between time, sex, or ethnicity
and the individual NEFA on any of the outcome measures.
Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
unadjusted models and `r cite_st('gee_adj')` for fully adjusted models.
Consistent with the GEE modeling, results from modeling all NEFA in a
multivariate environment using PLS showed no discernible clustering nor was it
predictive of the outcomes (data not shown).

>>>>

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta-cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from further GEE analyses. There were `r meds_use[1]` and
`r meds_use[2]` participants who used lipid and cholesterol modifying medications
at the baseline and 6-year visit, respectively. Medication use did not influence
NEFA composition, the outcomes, or the GEE models (data not shown).

`r cite_f('nefa')` shows the compositional distribution of NEFA in the study
participants. Four individual NEFA made up the vast majority 
(`r sum(nefa$pct)`%) of the total NEFA fraction. The largest contributors
were `r nefa$c[1]`, `r nefa$c[2]`, `r nefa$c[3]`, and `r nefa$c[4]`. Raw
concentration values are shown in `r cite_st('nefa')`. All individual NEFA as
well as the total fraction had correlations that ranged from weak to null
(r<0.3) with participant characteristics (see `r cite_sf('corr_conc')` and 
`r cite_sf("corr_pct")`).

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
There were several fatty acids (`r paste(unique(sig_fa_unadj$Xterms),collapse=", ")`)
that had significant associations with both insulin sensitivity and beta-cell
function outcomes. Most associations were negative, however, only 20:5n-3
(eicosapentaenoic acid; EPA) had positive associations with insulin sensitivity
as both a mol% and as a concentration. One fatty acid, 18:1n-7 (vaccenic acid),
had consistent negative associations with all four outcomes.

After full model adjustment (`r cite_f('gee_adj')`), most significant 
associations were attenuated, particularly with insulin sensitivity. The
remaining fatty acids (`r paste(unique(sig_fa_adj$Xterms),collapse=", ")`) had
significant associations, of which only EPA had a significant association (with
HOMA2-%S) as a mol% and was the only fatty acid with a positive association. All
other fatty acids had negative associations with IGI/IR and ISSI-2. The
magnitude of association for each of these variables was fairly consistent for
each beta-cell function measure. For every one SD increase in any of the NEFA
variables associated with beta-cell function, there was an average predicted
`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at any
given clinic visit.

Sensitivity analyses revealed that adjustment for WC was responsible for 
attenuating the unadjusted results (data not shown). Adjusting for TAG 
attenuated all associations with IGI/IR and ISSI-2 (data not shown), though this
may be due to the fact that only a subset of participants at the 6 year visit 
had TAG measured. Sensitivity analyses to examine the association with the sum 
of fatty acid classes (saturated vs unsaturated) revealed similar findings as 
seen with the adjusted GEE model for total NEFA (data not shown). There were no 
significant interaction effects of time, sex, or ethnicity in either unadjusted
or adjusted models. Medication use for controlling lipid and cholesterol did not
influence the results (data not shown). Raw values from the GEE models are shown
in `r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee_adj')` for
fully adjusted models. Consistent with the GEE modeling, results from modeling
all NEFA in a multivariate environment using PLS showed no discernible
clustering nor was it predictive of the outcomes (data not shown).
